### 2022-01-28 Paper.tex
modified

\section{Introduction}\label{sec:intro}
This supplement includes a description of the code, with a presentation of the features implemented followed by results of extensive benchmarks listed in
Table \ref{tab:list}, implemented to verify that the code solves correctly a large range of problems. Description and results of all benchmarks are shown
in Section \ref{sec:benchmark}.

\begin{table}[H]
\caption{List of benchmarks implemented for each feature of the code. Data of all benchmarks can be found at 
\url{https://github.com/aleregorda/Benchmarks}}
\centering
\begin{tabular}{| l | l | l |}
\toprule
 \textbf{Feature}  & \textbf{Benchmark} & \textbf{Section} \\
\midrule
  Solver  & Stokes flow & \ref{sec:stokes} \\
\hline
  4th-order Runge-Kutta  & \multirow{2}{*}{Zalesak disk} & \multirow{2}{*}{\ref{sec:runge}} \\
  markers advection  &  & \\
\hline
  \multirow{4}{*}{Momentum equation}  & Poiseuille flow & \ref{sec:poiseuille}\\
    & Instantaneous 2D sphere & \ref{sec:ist_sphere} \\
    & Rayleigh-Taylor experiment & \ref{sec:rayleigh} \\
    & Falling block & \ref{sec:block} \\
\hline
    & 2D Stokes sphere & \ref{sec:sphere} \\
    Sticky air and  & Stabilisation algorithm & \ref{sec:stab} \\
    free surface & Topography relaxation & \ref{sec:crameri} \\
  & Spontaneous subduction & \ref{sec:subduction} \\
\hline
  Erosion and sedimentation  & - & - \\
\hline
   Non-linear  & Slab detachment & \ref{sec:slab} \\
   visco-plasticity & Indenter & \ref{sec:indenter} \\
    & Brick & \ref{sec:brick} \\
\hline
  \multirow{3}{*}{Energy equation}  & Advection stabilisation & \ref{sec:advection} \\
    & Simple shear heating & \ref{sec:simple_shear} \\
    & Shear and adiabatic heating & \ref{sec:shear} \\
\hline
\multirow{3}{*}{Energy + momentum} & Mantle convection & \ref{sec:mantle} \\
& Viscoplastic mantle convection & \ref{sec:visco_mantle} \\
& Thin layer entrainment & \ref{sec:thin} \\
\hline
  Phase changes  & \multirow{2}{*}{Hydrated sinking cylinder} & \multirow{2}{*}{\ref{sec:quinquis}} \\
  and hydration  &  & \\
\hline
  Melting & Experimental melting curves & \ref{sec:katz} \\
\bottomrule
\end{tabular}
\label{tab:list}
\end{table}

@@

The time step $dt$ is chosen in according to Courant-Friedrichs-Lewy condition \citep{Anderson}, such that
\[dt=C_n\; \min\left(\frac{\Delta x}{\max|\bm{u}|},\frac{(\Delta x)^2}{\kappa}\right)\]
where $0<C_n<1$ is the Courant (CFL) number, $\Delta x$ is the minimum dimension of smallest element, $\max|\bm{u}|$ is the maximum velocity calculated on the
entire domain and
$\kappa=\frac{k}{\rho C_p}$ is the thermal diffusivity \citep{Thieulot2014}.

\section{Lagrangian markers}\label{sec:markers}
Elemental properties (density, viscosity, thermal conductivity, specific heat and thermal expansion) needed to solve Eqs. \ref{eq:mass_0}, \ref{eq:momentum}
and \ref{eq:energy} are related to the composition of each element, determined by means of Lagrangian markers that are characteristic of different materials
in the domain. Elemental properties (with the sole exception of the viscosity) are calculated using an arithmetic mean, as \[P_e=\frac{1}{n}\sum_{i=1}^n P_i\]
where $P_e$ is the elemental property, $P_i$ is the property characteristic of the material of each marker and $n$ is the number of markers of the element.
Differently, the average scheme for the elemental viscosity can be chosen between harmonic, geometric and arithmetic mean. Markers can be placed regularly or
randomly at the beginning of the simulation and their advection is performed by either a 2nd-order or a 4th-order Runge-Kutta in space, interpolating the velocity 
field on each marker by means of the shape functions (see Section \ref{sec:runge}). The interpolated velocity is then corrected by means of the Conservative
Velocity Interpolation (CVI), which introduces a corrective term to reduce dispersion and clustering of particles in both steady state and time-dependent 
\citep{Wang2015}.

@@

\section{Erosion and sedimentation}\label{sec:erosion}
Surface processes at the lithosphere-air interface are implemented taking into account fluvial erosion (long-range processes), hillslope processes (short-range
processes) and failure of slopes exceeding a threshold angle $\theta$ \citep{Burov1997,Tucker1999,Sacek2012,Croissant2014,Ueda2015,Beucher2020}. Erosion rates
$\dot{E}_{lr}$ due to long-range processes can be described by a stream power law equation \citep{Tucker1999,Ueda2015,Beucher2020}, in according to
\begin{equation}
\dot{E}_{lr}=K_f A^{m_e} \bm{\nabla} h^{n_e}
\end{equation}
where $K_f$ is the erodibility coefficient, $A$ is the upstream drainage area, $\bm{\nabla} h$ is the slope and $m_e$ and $n_e$ are parameters chosen in to
constrain the ratio $m_e/n_e$ between 0.35 and 0.8 \citep{Croissant2014,Ueda2015,Beucher2020}. Erosion rates $\dot{E}_{sr}$ due to short-range processes are
implemented by assuming that transport is linearly proportional to slope \citep{Burov1997,Ueda2015,Beucher2020}, in according to
\begin{equation}
\dot{E}_{sr}=K_d \bm{\nabla}^2 h
\end{equation}
where $K_d$ is the effective diffusivity. In addition, the slope is modified if it still exceeds a threshold angle after long- and short-range processes
\citep{Ueda2015}.

Offshore sedimentation processes are implemented similarly to hillslope processes but with a depth-dependent coefficient \citep{Sacek2012,Beucher2020}.
The sedimentation rate $\dot{S}$ is calculated as
\begin{equation}
\dot{S}=K_s \bm{\nabla}^2 h
\end{equation}
with
\begin{equation}
K_s=K^*_s \exp\left(\frac{h-h_{sl}}{h_0}\right)
\end{equation}
where $K^*_s$ is a constant, $h_{sl}$ is the sea level and $h_0$ is a characteristic length. In case of sticky air, slope is determined by means of the markers
chain, which is vertically corrected considering erosion and sedimentation rates. After the correction, continental markers above the markers chain are
transformed in air markers and air markers below the markers chain become sediments. Similarly, slope is determined considering the nodes on the top boundary
in case of true free surface and they are vertically corrected as for the markers chain. In this case, continental markers above the top boundary are simply
deleted.

\section{Non-linear rheologies}\label{sec:nonlinear}
Non-linear rheologies are implemented combining viscous creep (dislocation and diffusion) and plastic yielding. For each marker, diffusion and dislocation
viscosity ($\eta_{df|ds}$) can be determined as follows \citep{Karato1993,Wang2016}:
\begin{equation}\label{eq:viscous}
\eta_{df|ds}=\left(\frac{d^m}{A}\right)^{\frac{1}{n}}I_2^{\frac{(1-n)}{n}}\exp\left(\frac{Q+pV}{nRT}\right)
\end{equation}
where $d$ is the grain size, $m$ is the grain size exponent, $n$ is the stress exponent, $A$ is the pre-exponential factor, $I_2$ is the square root of the
second invariant of the strain rate tensor, $Q$ is the activation energy, $p$ is the pressure, $V$ is the activation volume, $R$ is the gas constant and $T$
is the temperature. Pressures and temperatures are determined by the interpolation of the nodal parameters, while strain rates are calculated by means of the
derivative of the velocities on the nodes. In case of diffusion creep $n=1$ and $m>0$, in case of dislocation creep $n>1$ and $m=0$. The implementation of the
non-linear viscous creep viscosity depending on the strain rate is tested by means of the slab detachment benchmark (Section \ref{sec:slab}).

Plastic yielding is implemented rescaling $\eta_{vc}$ in order to limit the stress below the yield stress $\sigma_y$ \citep{Thieulot2008,Thieulot2014,Glerum2018},
obtaining
\begin{equation}\label{eq:plastic}
\eta_{pl}=\frac{\sigma_y}{2I_2}
\end{equation}
where $\eta_{pl}$ is the plastic viscosity and the yield stress is determined following the Drucker-Prager criterion, such as
\begin{equation}\label{eq:yield}
\sigma_y=C\;\cos(\phi)+p\;\sin(\phi)
\end{equation}
where $C$ is the cohesion and $\phi$ is the internal friction angle. In case of negative pressure it is imposed equal to 0, so that negative yield stress are
excluded. The correctness of the non-linear solution in case of plastic viscosity with variable internal friction angle is verified performing indenter and
brick experiments (Sections \ref{sec:indenter} and \ref{sec:brick}, respectively).

Strain softening is taken into account for both viscous creep and plastic viscosity \citep{Huismans2003,Babeyko2005,Huismans2005,Sobolev2005,Warren2008a} by
means of the accumulated strain $\epsilon$ memorised by each marker. Softening in the viscous creep determines a linear decrease of $\eta_{vc}$ by means of a
viscous strain softening factor $W_S$ that increases linearly from 1 to 10 for $5<\epsilon<10$ \citep{Warren2008a}. This viscous softening can be related to
strain-induced grain size reduction \citep{Warren2008a}. Differently, plastic softening is simulated with a linear decrease of internal friction angle
$\phi(\epsilon)$ and cohesion $C(\epsilon)$ in according to
\begin{eqnarray}
\label{eq:friction1}\phi(\epsilon)&=&\phi_0+(\phi_{\infty}-\phi_0)\frac{\epsilon-\epsilon_0}{\epsilon_{\infty}-\epsilon_0}\\
\label{eq:friction2}C(\epsilon)&=&C_0+(C_{\infty}-C_0)\frac{\epsilon-\epsilon_0}{\epsilon_{\infty}-\epsilon_0}
\end{eqnarray}
where $\phi_0$, $C_0$ and $\phi_{\infty}$, $C_{\infty}$ are internal friction angle and cohesion for $\epsilon=0.5$ and $\epsilon=1.5$, respectively
\citep{Huismans2003,Huismans2005,Warren2008a,Thieulot2014}. Plastic softening approximates deformation-induced softening of faults and brittle shear zones
\citep{Warren2008a}.

Viscous creep and plastic yielding are then combined to obtain a viscoplastic viscosity $\eta_{vp}$, which is computed with an harmonic average
\citep{Tackley2010,Glerum2018} as follows
\begin{equation}\label{eq:viscoplastic}
\eta_{vp}=\left(\frac{1}{\eta_{df}}+\frac{1}{\eta_{ds}}+\frac{1}{\eta_{pl}}\right)^{-1}
\end{equation}
Finally, effective viscosity $\eta_{eff}$ is capped by the minimum and the maximum viscosity ($\eta_{min}$ and $\eta_{max}$, respectively) to avoid extremely
low or high viscosity \citep{Glerum2018} as follows
\begin{equation}\label{eq:effective}
\eta_{eff}=\min(\max(\eta_{vp},\eta_{min}),\eta_{max})
\end{equation}
Viscous creep and plastic yielding are non-linear rheologies because of their dependence on the velocity field through pressure and strain rates. Therefore,
the solution is determined by means of Picard-type iterations, until convergence of the velocity field \citep{Glerum2018}. The convergence is verified at each
iteration $i$ via the nonlinear residual $\bm{\mathcal{R}}^i$ that can be determined as
\begin{equation}\label{eq:convergence}
\bm{\mathcal{R}}^i= \mathbb{K} (\eta_{eff}(\dot{\bm{\epsilon}}^{i-1},p^{i-1}))\cdot \bm{v}^{i-1} - \bm{f}^i
\end{equation}
%\begin{equation}\label{eq:convergence}
%\chi=1-\frac{\langle (\bm{v}^i-\langle\bm{v}^i\rangle)\cdot (\bm{v}^{i+1}-\langle\bm{v}^{i+1}\rangle) \rangle}{\sqrt{\sigma^i\sigma^{i+1}}}
%\end{equation}
%Iterations are performed until either convergence ($\chi<tol$) or maximal number of non-linear iterations ($it_{max}$) is reached.
where $\mathbb{K}$ is the Stiffness matrix and $\bm{f}$ is the right hand side vector \citep{Spiegelman2016,Glerum2018}.  The $L_2$-norm is extracted from
$\bm{\mathcal{R}}^i$ and it is normalised as follows
\begin{equation}\label{eq:convergence_res}
\frac{||\bm{\mathcal{R}}^i||_2}{||\bm{\mathcal{R}}^0||_2}
\end{equation}
where $||\bm{\mathcal{R}}^0||_2$ is the $L_2$-norm of the first nonlinear iteration.
Since often the normalised nonlinear residual drop very quickly over the first 2-3 iterations, also the $L_2$-norm of the velocity residual is calculated as
\begin{equation}\label{eq:convergence_vel}
\frac{||\bm{v}_1-\bm{v}_{1-1}||_2}{||\bm{v}_{i-1}||_2}
\end{equation}
Iterations are performed until either maximal number of non-linear iterations ($it_{max}$) is reached or both the normalised nonlinear residual and the
velocity residual converge under a defined tolerance $tol$.

@@

\subsection{Markers advection}\label{sec:runge}
The 4th-order Runge-Kutta advection scheme in space is tested by means of the Zalesak disk test \citep{Zalesak1979,Thieulot2014}. The benchmark is performed
in a unit square domain with different grid resolution (\num{100x100} and \num{200x200} elements) and 1000000 markers. The Courant number varies between 0.1
and 0.5 and the velocity field is prescribed in the entire domain as
\begin{eqnarray}
u(x,y)&=&2\pi\left(y-\frac{L_x}{2}\right)\nonumber \\
v(x,y)&=&-2\pi\left(x-\frac{L_x}{2}\right)\nonumber
\end{eqnarray}
At $t=0$, the disk is centred at position $(0.5;0.75)$ with a radius $R=0.15$ and has a vertical fissure 0.05 wide and 0.2 high (orange in Fig. \ref{fig:runge}).
As we expect, markers are back to their initial location after a $2\pi$ rotation ($t=1$, red, green and blue in Fig. \ref{fig:runge}), with an error of the
final position that increase moving away from the centre, more evident in case of higher time steps or lower grid resolution (Fig. \ref{fig:runge_err}). 
All data can be found at \url{https://github.com/aleregorda/Benchmarks/tree/main/Zalesak_Disk}.

\begin{figure}
\centering
\noindent\includegraphics[width=200px]{./Figures/Zalesak.pdf}
\caption{Comparison between the initial position of markers of the Zalesak disk experiment (orange) and their final position for different CFL in case of grid
resolution of \num{100x100} elements. Blue, green and red markers indicate the final position of the disk in case of Courant number of 0.1, 0.25 and 0.5,
respectively.}
\label{fig:runge}
\end{figure}

\begin{figure}
\centering
\noindent\includegraphics[width=400px]{./Figures/zal_error.pdf}
\caption{Error between final and initial position of markers of the disk as function of their distance from the centre of the domain. Results for different
time steps are shown in blue, red and green for a grid resolution of \num{100x100} elements (for CFL of 0.1, 0.25 and 0.5, respectively) and in light blue,
dark green and purple in case of \num{200x200} elements (for CFL of 0.1, 0.25 and 0.5, respectively).}
\label{fig:runge_err}
\end{figure}

>>>>

\section{Introduction}\label{sec:intro}
This document includes a description of the code FALCON, with a presentation of the features implemented followed by results of extensive benchmarks listed in
Table \ref{tab:list}, performed to verify that the code solves correctly a wide range of problems. Description and results of all benchmarks are shown
in Section \ref{sec:benchmark}.

\begin{table}[H]
\caption{List of benchmarks performed for each feature of the code. Data of all benchmarks can be found at 
\url{https://github.com/aleregorda/Benchmarks}}
\centering
\begin{tabular}{| l | l | l |}
\toprule
 \textbf{Feature}  & \textbf{Benchmark} & \textbf{Section} \\
\midrule
  Solver  & Stokes flow & \ref{sec:stokes} \\
\hline
  \multirow{2}{*}{Markers advection} & Zalesak disk & \ref{sec:runge} \\
    & Conservative Interpolation Velocity & \ref{sec:CVI} \\
\hline
  \multirow{4}{*}{Momentum equation}  & Poiseuille flow & \ref{sec:poiseuille}\\
    & Instantaneous 2D sphere & \ref{sec:ist_sphere} \\
    & Rayleigh-Taylor experiment & \ref{sec:rayleigh} \\
    & Falling block & \ref{sec:block} \\
\hline
    & 2D Stokes sphere & \ref{sec:sphere} \\
    Sticky air and  & Stabilisation algorithm & \ref{sec:stab} \\
    free surface & Topography relaxation & \ref{sec:crameri} \\
  & Spontaneous subduction & \ref{sec:subduction} \\
\hline
  Erosion and sedimentation  & - & - \\
\hline
   Non-linear  & Slab detachment & \ref{sec:slab} \\
   visco-plasticity & Indenter & \ref{sec:indenter} \\
    & Brick & \ref{sec:brick} \\
\hline
  \multirow{3}{*}{Energy equation}  & Advection stabilisation & \ref{sec:advection} \\
    & Simple shear heating & \ref{sec:simple_shear} \\
    & Shear and adiabatic heating & \ref{sec:shear} \\
\hline
\multirow{3}{*}{Energy + momentum} & Mantle convection & \ref{sec:mantle} \\
& Viscoplastic mantle convection & \ref{sec:visco_mantle} \\
& Thin layer entrainment & \ref{sec:thin} \\
\hline
  Phase changes  & \multirow{2}{*}{Hydrated sinking cylinder} & \multirow{2}{*}{\ref{sec:quinquis}} \\
  and hydration  &  & \\
\hline
  Melting & Experimental melting curves & \ref{sec:katz} \\
\bottomrule
\end{tabular}
\label{tab:list}
\end{table}

@@

The time step $dt$ is chosen in according to Courant-Friedrichs-Lewy condition \citep{Anderson1995}, such that
\[dt=C_n\; \min\left(\frac{\Delta x}{\max|\bm{u}|},\frac{(\Delta x)^2}{\kappa}\right)\]
where $0<C_n<1$ is the Courant (CFL) number, $\Delta x$ is the minimum dimension of smallest element, $\max|\bm{u}|$ is the maximum velocity calculated on the
entire domain and
$\kappa=\frac{k}{\rho C_p}$ is the thermal diffusivity \citep{Thieulot2014}.

\section{Lagrangian markers}\label{sec:markers}
Elemental properties (density, viscosity, thermal conductivity, specific heat and thermal expansion) needed to solve Eqs. \ref{eq:mass_0}, \ref{eq:momentum}
and \ref{eq:energy} are related to the composition of each element, determined by means of Lagrangian markers that are characteristic of different materials
in the domain. Elemental properties (with the sole exception of the viscosity) are calculated using an arithmetic mean, as \[P_e=\frac{1}{n}\sum_{i=1}^n P_i\]
where $P_e$ is the elemental property, $P_i$ is the property characteristic of the material of each marker and $n$ is the number of markers of the element.
Differently, the average scheme for the elemental viscosity can be chosen between harmonic, geometric and arithmetic mean. Markers can be placed regularly or
randomly at the beginning of the simulation and their advection is performed by either a 2nd-order or a 4th-order Runge-Kutta in space, interpolating the velocity 
field on each marker by means of the shape functions (see Section \ref{sec:runge}). The interpolated velocity is then corrected by means of the Conservative
Velocity Interpolation (CVI), which introduces a corrective term to reduce dispersion and clustering of particles in both steady state and time-dependent 
\citep{Wang2015} (see Section \ref{sec:CVI}).

@@

\section{Erosion and sedimentation}\label{sec:erosion}
Surface processes at either the lithosphere-air interface or in correspondence of the free surface have been implemented by means of the software
Fastscape\footnote{\url{https://fastscape.org/}} \citep{Braun2013,Cordonnier2019,Yuan2019a,Yuan2019b}. FastScapeLib is a set of routines that solve the stream
power law (enriched by a sediment transport/deposition term) \citep{Yuan2019a}, hillslope diffusion and marine transport and deposition \citep{Yuan2019b}.
The partial differential equation solved by FastScapeLib is:
\begin{equation}
\frac{\partial h}{ \partial t}=U-K_f A^{m_e}S^{n_e} + \frac{G}{A}\int_A \left( U-\frac{\partial h}{\partial t} \right) dA + K_d \bm{\nabla}^2 h 
\end{equation}
where $h$ is the topography, $U$ is the uplift, $K_f$ is the erodibility coefficient, $A$ is the upstream drainage area, $S$ is the slope, $m_e$ and $n_e$ are
parameters that should be chosen to constrain the ratio $m_e/n_e$ between 0.35 and 0.8 \citep{Croissant2014,Ueda2015,Beucher2020}, $G$ is the dimensionless
deposition/transport coefficient for the enriched stream power law and $K_d$ is the effective diffusivity. Terms $K_f A^{m_e}S^{n_e}$ and $K_d \bm{\nabla}^2 h$
consider long-range processes, described by a stream power law equation \citep{Tucker1999,Ueda2015,Beucher2020}, and short-range processes,
proportional to slope \citep{Burov1997,Ueda2015,Beucher2020}.

In case of sticky air, slope is determined by means of the markers chain, which is vertically corrected considering erosion and sedimentation rates. After the
correction, continental markers above the markers chain are transformed in air markers and air markers below the markers chain become sediments. Similarly,
slope is determined considering the nodes on the top boundary in case of true free surface and they are vertically corrected as for the markers chain. In this
case, continental markers above the top boundary are simply deleted.
%Surface processes at the lithosphere-air interface are implemented taking into account fluvial erosion (long-range processes), hillslope processes (short-range
%processes) and failure of slopes exceeding a threshold angle $\theta$ \citep{Burov1997,Tucker1999,Sacek2012,Croissant2014,Ueda2015,Beucher2020}. Erosion rates
%$\dot{E}_{lr}$ due to long-range processes can be described by a stream power law equation \citep{Tucker1999,Ueda2015,Beucher2020}, in according to
%\begin{equation}
%\dot{E}_{lr}=K_f A^{m_e} \bm{\nabla} h^{n_e}
%\end{equation}
%where $K_f$ is the erodibility coefficient, $A$ is the upstream drainage area, $\bm{\nabla} h$ is the slope and $m_e$ and $n_e$ are parameters chosen in to
%constrain the ratio $m_e/n_e$ between 0.35 and 0.8 \citep{Croissant2014,Ueda2015,Beucher2020}. Erosion rates $\dot{E}_{sr}$ due to short-range processes are
%implemented by assuming that transport is linearly proportional to slope \citep{Burov1997,Ueda2015,Beucher2020}, in according to
%\begin{equation}
%\dot{E}_{sr}=K_d \bm{\nabla}^2 h
%\end{equation}
%where $K_d$ is the effective diffusivity. In addition, the slope is modified if it still exceeds a threshold angle after long- and short-range processes
%\citep{Ueda2015}.

%Offshore sedimentation processes are implemented similarly to hillslope processes but with a depth-dependent coefficient \citep{Sacek2012,Beucher2020}.
%The sedimentation rate $\dot{S}$ is calculated as
%\begin{equation}
%\dot{S}=K_s \bm{\nabla}^2 h
%\end{equation}
%with
%\begin{equation}
%K_s=K^*_s \exp\left(\frac{h-h_{sl}}{h_0}\right)
%\end{equation}
%where $K^*_s$ is a constant, $h_{sl}$ is the sea level and $h_0$ is a characteristic length. In case of sticky air, slope is determined by means of the markers
%chain, which is vertically corrected considering erosion and sedimentation rates. After the correction, continental markers above the markers chain are
%transformed in air markers and air markers below the markers chain become sediments. Similarly, slope is determined considering the nodes on the top boundary
%in case of true free surface and they are vertically corrected as for the markers chain. In this case, continental markers above the top boundary are simply
%deleted.

\section{Non-linear rheologies}\label{sec:nonlinear}
Non-linear rheologies are implemented combining viscous creep (dislocation and diffusion) and plastic yielding. For each marker, diffusion and dislocation
viscosity ($\eta_{df|ds}$) can be determined as follows \citep{Karato1993,Wang2016}:
\begin{equation}\label{eq:viscous}
\eta_{df|ds}=f_s \left(\frac{d^m}{A}\right)^{\frac{1}{n}} \left (\frac{2^{\frac{1-n}{n}}}{3^{\frac{n+1}{2n}}} \right)
I_2^{\frac{(1-n)}{n}}\exp\left(\frac{Q+pV}{nRT}\right)
\end{equation}
where $f_s$ is a scaling factor used to represent lithologies that are stronger or weaker than the base set, $d$ is the grain size, $m$ is the grain size exponent,
$n$ is the stress exponent, $A$ is the uniaxial pre-exponential factor, $I_2$ is the square root of the second invariant of the strain rate tensor, $Q$ is the
activation energy, $p$ is the pressure, $V$ is the activation volume, $R$ is the gas constant and $T$ is the temperature.
Pressures and temperatures are determined by the interpolation of the nodal parameters, while strain rates are calculated by means of the
derivative of the velocities on the nodes. In case of diffusion creep $n=1$ and $m>0$, in case of dislocation creep $n>1$ and $m=0$. The viscous creep $\eta_{cp}$
is then calculated as the harmonic average between $\eta_{df}$ and $\eta_{ds}$:
\begin{equation}\label{eq:visco_creep}
\eta_{cp}=\left(\frac{1}{\eta_{df}}+\frac{1}{\eta_{ds}}\right)^{-1}
\end{equation}
The implementation of the non-linear viscous creep viscosity depending on the strain rate is tested by means of the slab detachment benchmark
(Section \ref{sec:slab}).

Plastic yielding is implemented rescaling $\eta_{cp}$ in order to limit the stress below the yield stress $\sigma_y$ \citep{Thieulot2008,Thieulot2014,Glerum2018},
obtaining
\begin{equation}\label{eq:plastic}
\eta_{pl}=\frac{\sigma_y}{2I_2}
\end{equation}
where $\eta_{pl}$ is the plastic viscosity and the yield stress is determined following the Drucker-Prager criterion, such as
\begin{equation}\label{eq:yield}
\sigma_y=C\;\cos(\phi)+p\;\sin(\phi)
\end{equation}
where $C$ is the cohesion and $\phi$ is the internal friction angle. In case of negative pressure it is imposed equal to 0, so that negative yield stress are
excluded. The correctness of the non-linear solution in case of plastic viscosity with variable internal friction angle is verified performing indenter and
brick experiments (Sections \ref{sec:indenter} and \ref{sec:brick}, respectively).

Strain softening is taken into account for both viscous creep and plastic viscosity \citep{Huismans2003,Babeyko2005,Huismans2005,Sobolev2005,Warren2008a} by
means of the accumulated strain $\epsilon$ memorised by each marker. Softening in the viscous creep determines a linear decrease of $\eta_{vc}$ by means of a
viscous strain softening factor $W_S$ that increases linearly from $W_{S_0}$ to $W_{S_{\infty}}$ for $\epsilon_{S_0}<\epsilon_S<\epsilon_{S_{\infty}}$ 
\citep{Warren2008a}. This viscous softening can be related to strain-induced grain size reduction \citep{Warren2008a}. Differently, plastic softening is 
simulated with a linear decrease of internal friction angle $\phi(\epsilon)$ and cohesion $C(\epsilon)$ in according to
\begin{eqnarray}
\label{eq:friction1}\phi(\epsilon)&=&\phi_0+(\phi_{\infty}-\phi_0)\frac{\epsilon-\epsilon_0}{\epsilon_{\infty}-\epsilon_0}\\
\label{eq:friction2}C(\epsilon)&=&C_0+(C_{\infty}-C_0)\frac{\epsilon-\epsilon_0}{\epsilon_{\infty}-\epsilon_0}
\end{eqnarray}
where $\phi_0$, $C_0$ and $\phi_{\infty}$, $C_{\infty}$ are internal friction angle and cohesion for $\epsilon_0$ and $\epsilon_{\infty}$, respectively
\citep{Huismans2003,Huismans2005,Warren2008a,Thieulot2014}. Plastic softening approximates deformation-induced softening of faults and brittle shear zones
\citep{Warren2008a}.

Plastic yielding and viscous creep are then combined to obtain a viscoplastic viscosity $\eta_{vp}$ as follows, assuming that they are independent processes
\citep{Karato2008,Glerum2018}:
\begin{equation}\label{eq:viscoplastic}
\eta_{vp}=\min(\eta_{cp},\eta_{pl})
\end{equation}
Finally, effective viscosity $\eta_{eff}$ is capped by the minimum and the maximum viscosity ($\eta_{min}$ and $\eta_{max}$, respectively) to avoid extremely
low or high viscosity \citep{Glerum2018} as follows
\begin{equation}\label{eq:effective}
\eta_{eff}=\min(\max(\eta_{vp},\eta_{min}),\eta_{max})
\end{equation}
Viscous creep and plastic yielding are non-linear rheologies because of their dependence on the velocity field through pressure and strain rates. Therefore,
the solution is determined by means of Picard-type iterations, until convergence of the velocity field \citep{Glerum2018}. The convergence is verified at each
iteration $i$ via the nonlinear residual $\bm{\mathcal{R}}^i$ that can be determined as
\begin{equation}\label{eq:convergence}
\bm{\mathcal{R}}^i= \mathbb{K} (\eta_{eff}(\dot{\bm{\epsilon}}^{i-1},p^{i-1}))\cdot \bm{v}^{i-1} - \bm{f}^i
\end{equation}
%\begin{equation}\label{eq:convergence}
%\chi=1-\frac{\langle (\bm{v}^i-\langle\bm{v}^i\rangle)\cdot (\bm{v}^{i+1}-\langle\bm{v}^{i+1}\rangle) \rangle}{\sqrt{\sigma^i\sigma^{i+1}}}
%\end{equation}
%Iterations are performed until either convergence ($\chi<tol$) or maximal number of non-linear iterations ($it_{max}$) is reached.
where $\mathbb{K}$ is the Stiffness matrix and $\bm{f}$ is the right hand side vector \citep{Spiegelman2016,Glerum2018}.  The $L_2$-norm is extracted from
$\bm{\mathcal{R}}^i$ and it is normalised as follows
\begin{equation}\label{eq:convergence_res}
\frac{||\bm{\mathcal{R}}^i||_2}{||\bm{\mathcal{R}}^0||_2}
\end{equation}
where $||\bm{\mathcal{R}}^0||_2$ is the $L_2$-norm of the first nonlinear iteration.
Since often the normalised nonlinear residual drop very quickly over the first 2-3 iterations, also the $L_2$-norm of velocity and pressure residuals are
calculated as
\begin{equation}\label{eq:convergence_vel}
\frac{||\Delta\bm{v}||_2}{||\bm{v}||_2}=\frac{||\bm{v}_i-\bm{v}_{i-1}||_2}{||\bm{v}_{i}||_2}
\end{equation}
and
\begin{equation}\label{eq:convergence_press}
  \frac{||\Delta p||_2}{||p||_2}=\frac{||p_i-p_{i-1}||_2}{||p_{i}||_2}
\end{equation}
respectively. Iterations are performed until either maximal number of non-linear iterations ($it_{max}$) is reached or all the normalised nonlinear residual, 
the velocity residual and the pressure residual converge under a defined tolerance $tol$.

@@

\subsection{Markers advection}\label{sec:runge}
The 2nd-order Runge-Kutta advection scheme in space is tested by means of the Zalesak disk test \citep{Zalesak1979,Thieulot2014}. The benchmark is performed
in a unit square domain with a grid resolution of \num{32x32} elements and 1000000 markers and values of Courant number between 0.25 and 1.
At $t=0$, the disk is centred at position $(0.5;0.75)$ with a radius $R=0.15$ and has a vertical fissure 0.05 wide and 0.2 high (orange in Fig. \ref{fig:runge})
and with the velocity field prescribed in the entire domain as
\begin{eqnarray}
u(x,y)&=&2\pi\left(y-\frac{L_x}{2}\right)\nonumber \\
v(x,y)&=&-2\pi\left(x-\frac{L_x}{2}\right)\nonumber
\end{eqnarray}
the markers are back to their initial location after a $2\pi$ rotation ($t=1$). The clockwise rotation of the disk at $t=0.25$, 0.5 and 0.75 
is shown in Fig. \ref{fig:runge} (red, blue and green, respectively). The distance from the centre of 1 marker is calculated throughout the rotation to
evaluate the error for different values of Courant number. As expected, the error increase with the increasing of the time step, as shown in 
Fig. \ref{fig:runge_err}. 
All data can be found at \url{https://github.com/aleregorda/Benchmarks/tree/main/Advection/Zalesak_Disk}.

\begin{figure}
\centering
\noindent\includegraphics[width=200px]{./Figures/Zalesak.pdf}
\caption{Position of the Zalesak disk at $t=0$, 0.25, 0.5 and 0.75 (orange, red, blue and green, respectively) for the case with Courant number of 1.}
\label{fig:runge}
\end{figure}

\begin{figure}
\centering
\noindent\includegraphics[width=400px]{./Figures/zal_marker.pdf}
\caption{Distance from the centre as function of time for values of Courant number of  of 0.25, 0.3, 0.5, 0.75 and 1 (green, orange, blue, black and red,
respectively). R indicates the distance from the centre considering a perfect circle trajectory.}
\label{fig:runge_err}
\end{figure}

\subsection{Conservative Velocity Interpolation (CVI)}\label{sec:CVI}
The CVI correction is checked by means of the Stokes flow experiment by \citet{Donea2003}, as presented in \ref{sec:stokes}. The advection of Lagrangian
markers is performed using either a 2nd-order or a 4th-order Runge-Kutta scheme with CFL=0.5 and an initial random distribution of 25 markers per element.
Fig. \ref{fig:CVI} shows the minimum and maximum values of markers per element throughout the experiment with and without the CVI correction and using
different Runge-Kutta schemes. The introduction of the CVI correction allows the
tests to run for times longer than $t=2000$ without show empty elements, while tests without the CVI correction stop at approximately $t=600$, when at least
one element goes empty. Fig. \ref{fig:CVI_time} shows the different number of markers per element and the markers distribution at $t=600$ obtained without and
with the CVI correction (panels b and c, respectively) in case of a 2nd-order Runge-Kutta scheme. All data can be found at
\url{https://github.com/aleregorda/Benchmarks/tree/main/Advection/CVI}.
\begin{figure}
\centering
\noindent\includegraphics[width=\textwidth]{./Figures/CVI.pdf}
\caption{Maximum and minimum number of markers per element throughout the experiment, with (green and orange lines) or without (purple and blue lines)
the CVI correction and using either a 2nd-order (blue and orange lines) or a 4th-order (purple and green lines) Runge-Kutta scheme.}
\label{fig:CVI}
\end{figure}
\begin{figure}
\centering
\noindent\includegraphics[width=\textwidth]{./Figures/CVI_time.pdf}
\caption{Number of markers per element and markers distribution through time using 2nd.order Runge-Kutta scheme.
Panels a: number of markers per element (panel a1) and distribution of the markers (panel a2) after the first
iteration; panels b and c: comparison between number of markers per element (panels b1 and c1) and markers distribution (panels b2 and c2) without and with
the CVI correction at $t=600$; panels d: number of markers per element (panel d1) and markers distribution (panel d2) in case of CVI correction at $t=2000$.}
\label{fig:CVI_time}
\end{figure}
